An indexed construction of semi-simplicial and semi-cubical types

### 2022-05-19 paper/paper.tex
paper: minor backend changes; grouping commands
paper: figured out syntax for grouping; ready to merge into master
paper: roadblock in patching xkv; doesn't match original def

\usepackage{multirow, minted, tabularx, booktabs, setspace, makecell, caption}
\usepackage[hmargin=0.4in, vmargin=0.8in, headheight=10pt]{geometry}

@@

% Truncated sets, otherwise referred to as X: the general macro
\newcommandx{\X}[3][1=,2=,3=]{\ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{#2}{\setkeys*{X}{#3}}}}

% Frame, layer, and filler: the general macro
\newcommandx{\prim}[6][2=,3=,4=,5=,6=]{\ensuremath{\mathsf{\color{indian-yellow}{#1}}_{#2}^{\joinargs{}[#3][#4][#5]}{\setkeys*[prim]{#1}{#6}}}}

% Restriction for frame, layer, and filler: the general macro
\newcommandx{\restr}[8][2=,3=,4=,5=,6=,7=,8=]{\ensuremath{\mathsf{\color{russian-green}{restr}}_{\mathsf{\color{indian-yellow}{#1}}\joinargs{,}[#2][#3][#4]}^{\joinargs{}[#5][#6][#7]}{\setkeys*[restr]{#1}{#8}}}}

@@

% Hack: we'd like \coh to take 10 arguments, with the 10th being passed to
% xkeyval, but this is not possible. The only possible compromise is to join
% #7 and #8 by hand in usage of \coh: it's now #7 or '#6,#7' in \cohfoo commands
\newcommandx{\coh}[9][2=,3=,4=,5=,6=,7=,8=,9=]{\ensuremath{\mathsf{\color{chestnut}{coh}}_{\mathsf{\color{indian-yellow}{#1}}\joinargs{,}[#2][#3][#4][#5][#6]}^{\joinargs{}[#7][#8]}{\setkeys*[coh]{#1}{#9}}}}

@@

% The eqntable environment, displaying the various
\NewDocumentEnvironment{eqntable}{m}{\table[H]\small
\tabularx{\textwidth}{@{} l c c >{\centering\arraybackslash}X @{}}\toprule}
{\endtabularx\vspace{0.3em}\hrule\vspace{0.5em}\caption{#1}\endtable}

@@

\title{An indexed construction of semi-simplicial and semi-cubical types}
\author{
  \textcolor{gray80}{Hugo Herbelin} \\
  \itshape \textcolor{gray80}{IRIF}
  \and
  \textcolor{gray80}{Ramkumar Ramachandra} \\
  \itshape \textcolor{gray80}{Université de Paris}
}
\date{}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of \emph{homotopy type theory}, that formalizes \emph{truncated} $n$-types in dependent type theory. The accompanying artifact is an axiomless Coq formalization~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to truncated $n$-types; if one were to instantiate our development for a particular type (say, semi-simplicial types), this requirement vanishes.} that uses sophisticated rewriting techniques and pushes the boundary of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

\section{Simplicial sets}
Simplicial sets are powerful mathematical objects, that form the basis of much of modern homotopy theory. Their power derives from the fact that their homotopy category is exceptionally well-behaved, and that there are multiple model structures defined on them. Unlike CW-complexes that rely on pure topological notions like spheres and disks to ``tame'' a topological space, the definition of simplicial sets is purely categorical, and exist independent of topological spaces. The \emph{geometric realization} and \emph{singular complex} functors facilitate travelling back-and-forth between simplicial sets and topological spaces, and these are non-trivial constructions. As such, simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices.

\subsection{\texorpdfstring{\SSet}{The category of simplicial sets}}
There exists vast mathematical literature on the subject, and we refer the uninitiated reader to \cite{Friedman08} for a gentle introduction with excellent intutions. With these intuitions in place, we proceed with a purely categorical treatment. Emily Riehl's \cite{Riehl11} is an especially well-written manuscript on the subject.

\begin{definition}[\sq{n}]
  $\sq{n}$ is used to denote a totally-ordered set of natural numbers less than or equal to $n$, ordered by magnitude.

  \begin{equation*}
    \sq{n} := \{0, 1, \ldots, n\}
  \end{equation*}

  We will regard the totally-ordered set as a category in certain contexts, and it should be clear which interpretation we're referring to from the context.

  \begin{equation*}
    \sq{n} := 0 \rightarrow 1 \rightarrow \ldots \rightarrow n
  \end{equation*}
\end{definition}

\begin{definition}[\Simplex{}]
  The category $\Simplex{}$ is defined in terms of its objects and morphisms:

  \begin{align*}
    \obj(\Simplex{}) & := \sq{n}, \text{regarded as a totally-ordered set}   \\
    \mor(\Simplex{}) & := \sq{m} \rightarrow \sq{n}, \text{order-preserving}
  \end{align*}
\end{definition}

\begin{definition}[\SSet]
  The category of simplicial sets is defined as:

  \begin{equation*}
    \SSet := \Set^{\op{\Simplex{}}}
  \end{equation*}

  It is a functor category, whose objects and morphisms are given by:

  \begin{align*}
    \obj(\SSet)            & := \text{functors} \; \op{\Simplex{}} \rightarrow \Set     \\
    \mor(\SSet)            & := \text{natural transforms between the functors}          \\
    \text{composition law} & := \text{the usual composition of natural transformations} \\
  \end{align*}
\end{definition}

\begin{definition}[\Simplex{n}]
  The Yoneda embedding of $\sq{n}$ is referred to as the standard $n$-simplex.

  \begin{equation*}
    \Simplex{n} := y(\sq{n})
  \end{equation*}
\end{definition}

\begin{notation}[$\sq{0, \ldots, n}$]
  The notation $\sq{0, \ldots, n}$ denotes all ordered subsets of \sq{n}.
\end{notation}

With the definition of the category of simplicial sets firmly in place, we look inside a simplicial set to see the data it actually encodes concretely.

\begin{definition}[The data of a simplicial set]
  A simplicial set is equivalent to the data of a set of $n$-simplices, related to each other via face and degeneracy maps.

  \begin{align*}
    X_n                             & \;\;\text{a set of n-simplices} \\
    d_i : X_{n + 1} \rightarrow X_n & \;\;\text{face maps}            \\
    s_i : X_n \rightarrow X_{n + 1} & \;\;\text{degeneracy maps}      \\
  \end{align*}

  where $X_n$ is the set of $n$-simplices contained within the simplicial set $X$:

  \begin{align*}
    X_n & : \Set                                             \\
    X_n & := X(\sq{n}), \text{where X is the simplicial set} \\
  \end{align*}

  and $d_i$ and $s_i$ operate on $X_n$ as follows (the hat on $i$ indicates that it is to be omitted):

  \begin{align*}
    d_i(\sq{0, \ldots, n}) & = \sq{0, \ldots, \hat{i}, \ldots, n} \\
    s_i(\sq{0, \ldots, n}) & = \sq{0, \ldots, i, i, \ldots, n}    \\
  \end{align*}

  and $d_i$, $s_i$ are constrained by the following ``simplicial identities'':

  \begin{equation*}
    \begin{matrix}
      d_i d_j & = & d_{j - 1} d_i & i < j \\
      s_i s_j & = & s_{j + 1} s_i & i < j \\
      d_i s_j & = &
      \begin{cases}
        s_{j - 1} d_i & i < j \\
        id            & i = j \\
        s_j d_{i - 1} & i > j \\
      \end{cases}
    \end{matrix}
  \end{equation*}
\end{definition}

We mentioned degeneracies in the introduction to this section, and we define precisely what a degenerate simplex is now.

\begin{definition}[Degenerate simplex]
  A simplex is $x \in X_n$ is termed degenerate if it is the image of some degeneracy map $s_i$, and non-degenerate otherwise.
\end{definition}

\begin{example}[Non-degenerate simplices of \Simplex{n}]
  The non-degenerate $k$-simplices of $\Simplex{n}$ are the injective maps $\sq{k} \rightarrow \sq{n}$ in $\Simplex{}$. In particular, $\Simplex{n}$ has a unique non-degenerate $n$-simplex.
\end{example}

As is clear from the above definitions, simplicial sets have an infinite number of degeneracies, and the first step towards drawing a simplicial set is to define a convention that omits these degeneracies in the drawing.

\begin{notation}[Drawing of a simplicial set\label{not:drawsset}]
  A diagram of the form

  \begin{equation*}
    \begin{tikzcd}
      x \arrow[r, "f"] & y
    \end{tikzcd}
  \end{equation*}

  in simplicial set $S$ will mean that $f$ is a non-degenerate $1$-simplex, and $x$ and $y$ are $0$-simplices. The simplices have a relationship to the simplicial set $S$ given by the following commutative diagram:

  \begin{equation*}
    \begin{tikzcd}
      \Simplex{0} \arrow[dr] \arrow[drr, "x", bend left] & & \\
      & \Simplex{1} \arrow[r, "f"] & S \\
      \Simplex{0} \arrow[ur] \arrow[urr, "y"', bend right] & & \\
    \end{tikzcd}
  \end{equation*}

  To attempting to draw simplices of dimension greater than $1$, we see that there is a potential ambiguity. Consider the following diagram:

  \begin{equation*}
    \begin{tikzcd}
      & \bullet \arrow[ddr] & \\
      \\
      \bullet \arrow[uur] \arrow[rr] & & \bullet
    \end{tikzcd}
  \end{equation*}

  In the above diagram, it is clear that there are three non-degenerate $1$-simplices, but it is unclear whether there is a non-degenerate $2$-simplex. By convention, when we draw such a diagram, we will assume that there is a non-degenerate $2$-simplex border by the three non-degenerate $1$-simplices. To generalize, we will always assume that there exists a non-degenerate $n$-simplex, when bounded by non-degenerate $(n - 1)$-simplices.
\end{notation}

\begin{example}[\Simplex{0}, \Simplex{1}, and \Simplex{2}]
  $\Simplex{0}$ can be drawn as:

  $$
    \begin{tikzcd}
      0
    \end{tikzcd}
  $$

  $\Simplex{1}$ can be drawn as:

  $$
    \begin{tikzcd}
      0 \arrow[r] & 1
    \end{tikzcd}
  $$

  and $\Simplex{2}$ can be drawn as:

  $$
    \begin{tikzcd}
      & 1 \arrow[ddr] & \\
      \\
      0 \arrow[uur] \arrow[rr] & & 2
    \end{tikzcd}
  $$
\end{example}

\subsection{\texorpdfstring{\DeltaHat}{Delta sets}}
The first step towards formalizing simplicial sets is to forget about the degeneracies, and we restrict ourselves to representing \emph{augmented simplicial sets}, otherwise known as delta-sets in mathematical literature. We now proceed to describe \DeltaHat, which is identical to \Simplex{}, but for the fact that the maps $\sq{m} \rightarrow \sq{n}$ are \emph{strictly} order-preserving. The degeneracies then vanish, and we're left with the task of defining gluing conditions solely on the basis of face maps. In mathematical literature, there is little interest in studying \DeltaHat, since maps between delta sets are not well-defined, and it is therefore not possible to construct a category of delta sets.

@@

\section{Grothendieck construction\label{sec:grothendieckconstr}}
The Grothendieck construction provides a correspondence between a fibered representation, and an indexed representation.

@@

\section{Unicity of identity proofs\label{sec:uip}}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

@@

\section{Trunacated \texorpdfstring{$n$}{n}-sets}
In this section, we give the core of the definition of truncated $n$-sets
as a coinductive limit.

@@

% For just the first table, keep the universe letter in fullframe
\newcommandx{\fullframe}[3][1=,2=,3=]{\prim{fullframe}[#1][#2][][][#3]}
\input{tab-core.tex}

@@

@@

% Truncated sets, otherwise referred to as X
% Keep universe letter for last table
\renewcommandx{\Xto}[2]{\X[#1][<#2]}
\renewcommandx{\Xcomp}[2]{\X[#1][=#2]}
\renewcommandx{\Xfrom}[2]{\X[#1][\geq#2]}

@@

\input{tab-core.tex}
\input{tab-frames.tex}
\input{tab-faces.tex}
\input{tab-coh.tex}
\input{tab-coind.tex}

>>>>

\usepackage{multirow, minted, tabularx, booktabs, setspace, makecell, caption, xstring, pdflscape}
\usepackage[a4paper, inner=25mm, outer=25mm, top=25mm, bottom=25mm]{geometry}

@@

% More magic: we extend xkeyval with a \appendmask macro to pass to \setkeys
\makeatletter
\edef\bonak@mask{}

\def\bonak@appendmask[#1]{
  \XKV@addtolist@n\bonak@mask{\g@addto@macro{\XKV@prefix\XKV@tfam}{@}#1}
}
\def\appendmask{\XKV@testopta{\XKV@testoptc\bonak@appendmask}}
\makeatother

% Now we patch XKV@setkeys
\makeatletter
\let\XKV@setkeys@alt\XKV@setkeys
\long\def\XKV@setkeys[#1]#2{
  \edef\bonak@pfx{\g@addto@macro{\XKV@prefix\XKV@tfam}{@}}
  \edef\bonak@tempa\@empty
  \edef\bonak@keys{}
  \XKV@for@n{\bonak@mask}\bonak@tempa{
    \IfBeginWith{\bonak@tempa}{\bonak@pfx}{
      \XKV@addtolist@n\bonak@keys{\StrBehind{\bonak@tempa}{\bonak@pfx}}}{}
  }
  % Finally, call the real function
  \XKV@setkeys@alt[\bonak@keys]{#2}
}
\makeatother

% Truncated sets, otherwise referred to as X: the general macro
\newcommandx*{\X}[3][1=,2=,3=]{
  \ensuremath{\mathsf{\color{carolina}{X}}_{#1}^{#2}}
  \setkeys*{X}{#3}
}

% Frame, layer, and filler: the general macro
\newcommandx*{\prim}[6][2=,3=,4=,5=,6=]{
  \ensuremath{\mathsf{\color{indian-yellow}{#1}}_{#2}^{
    \joinargs{}[#3][#4][#5]}}
  \setkeys*[prim]{#1}{#6}
}

% Restriction for frame, layer, and filler: the general macro
\newcommandx*{\restr}[8][2=,3=,4=,5=,6=,7=,8=]{
  \ensuremath{\mathsf{\color{russian-green}{restr}}_{
    \mathsf{\color{indian-yellow}{#1}}
    \joinargs{,}[#2][#3][#4]}^{\joinargs{}[#5][#6][#7]}}
  \setkeys*[restr]{#1}{#8}
}

@@

% Hack: we'd like \coh to take 10 arguments, with the 10th being passed to
% xkeyval, but this is not possible. The only possible compromise is to join
% #7 and #8 by hand in usage of \coh: it's now #7 or '#6,#7' in \cohfoo commands
\newcommandx*{\coh}[9][2=,3=,4=,5=,6=,7=,8=,9=]{
  \ensuremath{\mathsf{\color{chestnut}{coh}}_{
    \mathsf{\color{indian-yellow}{#1}}
    \joinargs{,}[#2][#3][#4][#5][#6]}^{\joinargs{}[#7][#8]}}
  \setkeys*[coh]{#1}{#9}
}

@@

% The eqntable environment, displaying the various
\NewDocumentEnvironment{eqntable}{m}{\table[H]\small\hrule\vspace{0.5em}
\tabularx{\textwidth}{@{} l c c >{\centering\arraybackslash}X @{}}}
{\endtabularx\vspace{0.3em}\hrule\vspace{0.5em}\caption{#1}\endtable}

@@

\title{An indexed construction of semi-simplicial and semi-cubical types}
\author{
  \textcolor{gray80}{Ramkumar Ramachandra} \\
  \itshape \textcolor{gray80}{Université Paris-Cité}
  \and
  \textcolor{gray80}{Hugo Herbelin} \\
  \itshape \textcolor{gray80}{Instit de Recherche en Informatique Fondamentale}
}
\date{}

@@

\begin{document}
\thispagestyle{empty}
\maketitle
\begin{abstract}
  A model of homotopy type theory, that formalizes augmented semi-simplicial sets that naturally generalizes to semi-cubical sets as well, in dependent type theory. The accompanying artifact is an axiomless~\footnote{Technically speaking, we use some axioms related to functional extensionality, but the requirement arises only from our generalization to $\nu$-types; if $\nu$ is finite (say, unit type corresponding to augmented semi-simplicial types), this requirement vanishes.} Coq formalization that uses sophisticated rewriting techniques and pushes the boundaries of proof assistant technology.
\end{abstract}
\tableofcontents
\newpage

@@

\section{\texorpdfstring{$\nu$}{nu}-sets}
Simplicial sets are powerful mathematical objects, that form the basis of much of modern homotopy theory. Their power derives from the fact that their homotopy category is exceptionally well-behaved, and that there are multiple model structures defined on them. Unlike CW-complexes that rely on pure topological notions like spheres and disks to ``tame'' a topological space, the definition of simplicial sets is purely categorical, and exist independent of topological spaces. The \emph{geometric realization} and \emph{singular complex} functors facilitate travelling back-and-forth between simplicial sets and topological spaces, and these are non-trivial constructions. As such, simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices.

\subsection{Augmented semi-simplicial sets}
The first step towards formalizing simplicial sets is to forget about the degeneracies, and we restrict ourselves to representing \emph{augmented simplicial sets}, otherwise known as delta-sets in mathematical literature. We now proceed to describe \DeltaHat, which is identical to \Simplex{}, but for the fact that the maps $\sq{m} \rightarrow \sq{n}$ are \emph{strictly} order-preserving. The degeneracies then vanish, and we're left with the task of defining gluing conditions solely on the basis of face maps. In mathematical literature, there is little interest in studying \DeltaHat, since maps between delta sets are not well-defined, and it is therefore not possible to construct a category of delta sets.

@@

\section{Our construction}
The Grothendieck construction provides a correspondence between a fibered representation, and an indexed representation.

@@

\section{Reasoning about sets in type theory}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

@@

\section{Type-theoretic construction of \texorpdfstring{$\nu$}{nu}-sets}
In this section, we give the core of the definition of truncated $\nu$-sets
as a coinductive limit.

@@

% For just the second table, keep the universe letter in fullframe
\newcommandx{\fullframe}[3][1=,2=,3=]{\prim{fullframe}[#1][#2][][][#3]}
\input{tab-core.tex}

@@

% From the third table onwards, make the {D, E} arguments of prims implicit
\appendmask[prim]{frame}[D]
\appendmask[prim]{layer}[D]
\appendmask[prim]{filler}[D, E]

@@

@@

\input{tab-coind.tex}
\input{tab-core.tex}
\input{tab-frames.tex}

\begin{landscape}
  \input{tab-faces.tex}
  \input{tab-coh.tex}
\end{landscape}
